---
title: "Rproject_semesterproj"
author: "Lisa Wahlen"
date: "`r Sys.Date()`"
output: html_document
editor_options: 
  markdown: 
    wrap: 72
---

### Libraries

```{r}

# had to install Rtools43 before i could run the code below : https://cran.rstudio.com/bin/windows/Rtools/rtools43/rtools.html

# devtools::install_github("benmarwick/wordcountaddin",  type = "source", dependencies = TRUE)
# wordcountaddin:::text_stats() # to check character count
# install.packages("koRpus")

library(wordcountaddin)
library(koRpus)

wordcountaddin:::text_stats()
```


```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)


library(readr) # to import tabular data (e.g. csv)
library(dplyr) # to manipulate (tabular) data
library(ggplot2) # to visualize data
library(sf) # to handle spatial vector data
library(terra) # To handle raster data
library(lubridate) # To handle dates and times
library(SimilarityMeasures)
library(RColorBrewer)
library(tmap)


mycolors4 <- brewer.pal(4, "BrBG")
# BrBG palette: #a6611a, #dfc27d, #80cdc1, #018571
theme_set(theme_light())
```

### Data

```{r}
posmo <- read_csv("data/posmo_2023-05-18T00_00_00+02_00-2023-05-18T23_59_59+02_00.csv")
```

Evtl. könnte man die Daten auf die AV-Daten legen. So kann man dann alle
Trajectories löschen, welche sich auf einer Strasse befinden.

```{r}
posmo <- st_as_sf(posmo, coords = c("lon_x","lat_y"), crs = 4326) |>
  st_transform(2056)

head(posmo)

posmo_coordinates <- st_coordinates(posmo)

posmo <- cbind(posmo, posmo_coordinates)

head(posmo) # there is an outlier. one point was recorded the night before the tour started.
# remove outlier
posmo <- posmo |> 
  slice(-1)

posmo_transport <- select(posmo, -c("weekday", "place_name", "user_id"))
posmo <- select(posmo, -c("weekday", "place_name", "transport_mode", "user_id"))
```

Dataset for verification
wiriehorn
```{r}
wiriehorn <- read_csv("data/posmo/mtb_wiriehorn.csv")
head(wiriehorn)
wiriehorn <- st_as_sf(wiriehorn, coords = c("lon_x","lat_y"), crs = 4326) |>
  st_transform(2056)

wiriehorn_coordinates <- st_coordinates(wiriehorn)

wiriehorn <- cbind(wiriehorn, wiriehorn_coordinates)
head(wiriehorn)

# again there is an outlier
wiriehorn <- wiriehorn |> 
  slice(-1)

wiriehorn_transport <- select(wiriehorn, -c("weekday", "place_name", "user_id"))

tmap_mode("view")
tm_shape(wiriehorn_transport)+
  tm_dots("transport_mode")+
  tm_layout(title= 'Wiriehorn trajectory of the 6th of May', title.position = c('right', 'top'))

# here we see that some parts of the trajectory were labelled as bike
```

take a look at the other datasets
valbirse
```{r}
valbirse <- read_csv("data/posmo/mtb_valbirse.csv")
head(valbirse)
valbirse <- st_as_sf(valbirse, coords = c("lon_x","lat_y"), crs = 4326) |>
  st_transform(2056)

valbirse_coordinates <- st_coordinates(valbirse)

valbirse <- cbind(valbirse, valbirse_coordinates)
head(valbirse)

# again there is an outlier
valbirse <- valbirse |> 
  slice(-1)

valbirse_transport <- select(valbirse, -c("weekday", "place_name", "user_id"))

tmap_mode("view")
tm_shape(valbirse_transport)+
  tm_dots("transport_mode")

ggplot(valbirse, aes(x=X, y=Y))+
  geom_path(aes(col=datetime))+
  coord_fixed()
```

solothurn

```{r}
solothurn <- read_csv("data/posmo/mtb_solothurn.csv")
head(solothurn)
solothurn <- st_as_sf(solothurn, coords = c("lon_x","lat_y"), crs = 4326) |>
  st_transform(2056)

solothurn_coordinates <- st_coordinates(solothurn)

solothurn <- cbind(solothurn, solothurn_coordinates)
head(solothurn)

solothurn_transport <- select(solothurn, -c("weekday", "place_name", "user_id"))

tmap_mode("view")
tm_shape(solothurn_transport)+
  tm_dots("transport_mode")
```

miniride

```{r}
miniride <- read_csv("data/posmo/mtb_miniride.csv")
head(miniride)
miniride <- st_as_sf(miniride, coords = c("lon_x","lat_y"), crs = 4326) |>
  st_transform(2056)

miniride_coordinates <- st_coordinates(miniride)

miniride <- cbind(miniride, miniride_coordinates)
head(miniride)

# again there is an outlier
miniride <- miniride |> 
  slice(-1)

miniride_transport <- select(miniride, -c("weekday", "place_name", "user_id"))

tmap_mode("view")
tm_shape(miniride_transport)+
  tm_dots("transport_mode")
```


### Visualizations

```{r}
# overview datapoints
ggplot(posmo, aes(x=X, y=Y))+
  geom_point()+
  coord_fixed()

# overview transport mode
ggplot(posmo_transport, aes(x=X, y=Y, color=transport_mode))+
  geom_point()+
  coord_fixed()

# overview with map background
tmap_mode("view")
tm_shape(posmo_transport)+
  tm_dots("transport_mode")

```

most of the trajectory is labelled with the transport mode "car". The
loop located in the south of the trajectory shows the part where Lisa
went downhill-biking in Diemtigtal, Allmiried. Most of that section is
labelled as "car", wherease some points are labelled as "walk". We now
want to try to find out some criteria in order to be able to isolate
that downhill biking part and label it as "downhill biking"

## Isolation of mountainbiking pattern

### Criteria I: speed

Speed threshold compute movement parameter speed for every fix and a
profile of the movement parameter. (Dodge et al., 2009)

```{r}
# calculate distance between points d_plusx for distance
# calculate time between points
# calculate speed
# calculate average speed between four fixes

posmo_transport <- posmo_transport |> 
  mutate(
    d_plus1 = sqrt((lead(X)-X)^2 + (lead(Y)-Y)^2),
    d_plus2 = sqrt((lead(X,2)-X)^2 + (lead(Y,2)-Y)^2), # new offset of 2 in function lead()
    d_minus1 = sqrt((lag(X)-X)^2 + (lag(Y)-Y)^2),
    d_minus2 = sqrt((lag(X,2)-X)^2 + (lag(Y,2)-Y)^2),
    t_plus1 = difftime(lead(datetime), datetime),
    t_plus2 = difftime(lead(datetime, 2), datetime),
    t_minus1 = difftime(lag(datetime), datetime),
    t_minus2 = difftime(lag(datetime, 2), datetime),
    s_plus1 = d_plus1/as.numeric(t_plus1), # s for speed [m/s]
    s_plus2 = d_plus2/as.numeric(t_plus2),
    s_minus1 = d_minus1/-as.numeric(t_minus1),
    s_minus2 = d_minus2/-as.numeric(t_minus2)
  )

# calculate mean speed within window of two neighboring points. not exactly a temporal window because the sampling time varies slightly. mean in m/s and km/h.

posmo_transport <- posmo_transport |>
  rowwise() |>  # so the mean per row is computed
  mutate(
    speedmean_ms = mean(c(s_plus1, s_plus2, s_minus1, s_minus2), na.rm=TRUE)
  ) |> 
  ungroup() |> # otherwise each row would represent a group, we don't want that.
  # now  we calculated the mean speed in m/s.
  # calculate mean speed in km per hour. 1 m/s = 3.6 km/h
  mutate(speedmean_kmh = speedmean_ms * 3.6)


```


#### visualisation of average speed, dealing with NaN

```{r}
# explore what we just did
ggplot(posmo_transport, aes(speedmean_ms))+
  geom_histogram(binwidth = 1)+
  geom_vline(xintercept = mean(posmo_transport$speedmean_ms, na.rm=TRUE)) # why is there no vline?

ggplot(posmo_transport, aes(speedmean_kmh))+
  geom_histogram(binwidth = 1)+
  geom_vline(xintercept = mean(posmo_transport$speedmean_kmh, na.rm=TRUE)) # still no vline?


range(posmo_transport$speedmean_ms, na.rm=TRUE)

posmo_transport$speedmean_ms[1:50] # how did i get inf and NaN values? something divided by 0 returns inf. 0/0 returns NaN. no vline because of nan values
pt <- is.finite(posmo_transport$speedmean_ms) # returns TRUE if it is not one of the values NA, NaN, Inf or -Inf
length(pt[pt == FALSE]) # counts NA, NaN, Inf and -Inf. we have 80 of those

# plot points with inf and non numeric value. let's see which points are affected.

inf_filter <- posmo_transport |> 
  filter(!is.finite(speedmean_ms))

tmap_mode("view")
tm_shape(posmo_transport)+
  tm_dots()+
  tm_shape(inf_filter)+
  tm_dots("transport_mode")

# of 2,237 rows only 80 rows are affected (3.58 %). since the +-Inf, NaN and NA values emerge where either the distance or the time between two points is 0, the points in question seem to be static (either temporally or locally). where no speed can be computed, no evaluation regarding a threshold can be taken. 
# the inf/NaN/NA points are located where little movement is plausible (at the bottom and the top of the cable car allmiried, somewhere in the middle of the downhill trail, at train stations in Solothurn). 

# new dataset to create segmentation
posmo_segments <- posmo_transport |> 
    mutate(static = !is.finite(posmo_transport$speedmean_ms)) # returns TRUE for static points

# "clean" dataset for visualisation of moving point parameter profile
posmo_transport <- posmo_transport |>
  filter(is.finite(speedmean_ms))

ggplot(posmo_transport, aes(speedmean_ms))+
  geom_histogram(binwidth = 1)+
  geom_vline(xintercept = mean(posmo_transport$speedmean_ms)) # vline is visible, average speed on that day was around 10 m/s

ggplot(posmo_transport, aes(speedmean_kmh))+
  geom_histogram(binwidth = 1)+
  geom_vline(xintercept = mean(posmo_transport$speedmean_kmh)) # vline is visible, average speed on that day was around 37.33548 km/h

mean(posmo_transport$speedmean_kmh)
```

#### moving parameter (MP) profile for speed (Dodge et
al., 2009)

```{r}
head(posmo_transport)

ggplot(posmo_transport, aes(x = datetime, y = speedmean_kmh))+
  geom_line()+
  geom_point(col ="#dfc27d", alpha = .2)+
  ylab("average speed [km/h]")+
  xlab("time")+
  labs(title = "Movement parameter profile for the average speed")

```

**Looking at the plot above, we can see, that the mountainbiking pattern
must have been taking place between 9 am and about 1.30 pm. Therefore we
will look a bit closer at the exact speed during this time**. to set a reasonable
speed range or threshold, i want to figure out the average speed of a
bike descent. i try to isolate the bike segments.

Segments visualisation
where are the static points in the allmiried area? i see static points at the talstation, bergstation and halfway on the trail. (breaks to wait for each other, slow movements waiting for the funiculare)
```{r}
tmap_mode("view")
tm_shape(posmo_segments)+
  tm_dots("static")
```

#### segmentation
assign unique IDs to the segments

```{r}
# function for assigning unique ID to the subtrajectories

rle_id <- function(vec) {
    x <- rle(vec)$lengths
    as.factor(rep(seq_along(x), times = x))
}

posmo_segments <- posmo_segments |> 
  mutate(ID=rle_id(static))

```

Visualize the moving segments by colourizing them by segment_ID.

```{r}

posmo_segments <- posmo_segments |> 
  filter(!static) # select rows where the static value is not TRUE to select the moving segments

ggplot(posmo_segments, aes(X, Y))+
  geom_point()+
  geom_path(aes(color=ID))+
  coord_fixed()

tm_shape(posmo_segments)+
  tm_dots("ID")+
  tmap_options(max.categories = 37)

# MP profile
ggplot(posmo_segments, aes(x = datetime, y = speedmean_kmh))+
  geom_line(aes(color = ID))+
  ylab("average speed [km/h]")+
  xlab("time")+
  labs(title = "Movement parameter profile of average speed", subtitle = "for moving segments of Allmiried tour")


```

#### choosing segments and defining range
looks like the segments 12, 14, 16, 18, 20, 22, 24, 26 are mountainbiking segments
selecting the segments between 9:00 and 13:00
```{r}
posmo_threshold <- posmo_segments |> 
  filter(ID == c(12, 14, 16, 18, 20, 22, 24, 26))

ggplot(posmo_threshold, aes(X, Y))+
  geom_point(aes(color=ID))+
  geom_path(aes(color=ID))+
  coord_fixed()

# covers the time range between 08:42 - 13:13 which corresponds to biking
# figure out which speed range makes sense to define mountainbiking

ggplot(posmo_threshold, aes(y=speedmean_kmh))+
  geom_boxplot()

summary(posmo_threshold$speedmean_kmh)
summary(posmo_threshold$speedmean_ms)
sd(posmo_threshold$speedmean_kmh)
```
speed summary for kmh
    Min.  1st Qu.   Median     Mean  3rd Qu.     Max. 
 0.08902  2.85510  8.26566  7.80428 10.30834 26.95490 

speed summary for ms
   Min. 1st Qu.  Median    Mean 3rd Qu.    Max. 
0.02473 0.79308 2.29602 2.16786 2.86343 7.48747 

according to the speed statistics i will choose the range of the 1st and 3rd quartile kmh (2.855 to 10.308) as the speed defining mountainbiking.

rename / overwrite rows of points within the speed range of mountainbiking.
```{r}

# posmo_transport$transport_mode[posmo_transport$speedmean_kmh >= 2.855 & posmo_transport$speedmean_kmh <= 10.308] <- "mtb"
# 
# tm_shape(posmo_transport)+
#   tm_dots("transport_mode")

# zooming into the allmiried area we can see that the cable car is labelled as mtb. the speed range therefore applies mainly to the cable car. let's see if by adjusting the range to include lower values as the set threshold changes anything.

# posmo_transport$transport_mode[posmo_transport$speedmean_kmh <= 10.308] <- "mtb"
# 
# tm_shape(posmo_transport)+
#   tm_dots("transport_mode")

# a few more points are included, a great deal of the downhill parts are still not covered though. adjust upper threshold, choose max value of statistics output above. exclude range of cable car

posmo_transport$transport_mode[posmo_transport$speedmean_kmh <= 26.95490 & !(posmo_transport$speedmean_kmh >= 2.855 & posmo_transport$speedmean_kmh <= 10.308)] <- "mtb"

tm_shape(posmo_transport)+
  tm_dots("transport_mode")

# zooming in the map, we see that the cable car is excluded in the allmiried area. since the walking speed of 4-5kmh is also excluded, the transport mode "walking" is preserved. however, some points on the mountainbike descent are excluded, probably because they fall in the speed range of the cable car. otherwise, the result represents most of the downhill mountainbiking.

ggplot(posmo_threshold, aes(x = datetime, y = speedmean_kmh))+
  geom_line()+
  geom_point(col ="#dfc27d", alpha = .7)

```
choosing the average speeds including the full range of the mountainbiking speeds did not exclude the cable car. taking a closer look a the moving parameter profile of the mountainbike segments we see that the cable car travels probably at an average of 8kmh. since average walking speed 

#### assigning transport mode mtb
now i have to make it compatible with the other dataset with the ground cover. there, mountainbiking is labelled in a separate column with the criteria "yes".
```{r}
posmo_transport <- posmo_transport |> 
  mutate(mountainbike = transport_mode =="mtb")

length(posmo_transport$mountainbike[posmo_transport$mountainbike == TRUE]) # 605 true

posmo_transport$mountainbike[posmo_transport$mountainbike == TRUE] <- "yes"

tm_shape(posmo_transport)+
  tm_dots("transport_mode")
```


### Criteria II: Ground cover

in a second step, we will try to do the according filtering but by using
a classification according to the DEM and to ground cover.

```{r}
#csv has all download paths from alti3d tiles:  https://www.swisstopo.admin.ch/en/geodata/height/alti3d.html 
# all_tif <- read.csv("data/alti3D_all.csv", header = FALSE)
# 
# # terra help https://rspatial.org/spatial-terra/8-rastermanip.html 
# 
# # download all files to an empty folder
# for (fi in all_tif$V1){
# outfile <- basename(fi)
# print(outfile)
# download.file(fi, paste0("data/alti3d_05/", outfile),mode = "wb") # mode (binary) very important
# }

# list takes all files with .tif from a folder
file_list <- list.files("data/alti3d_05/",".tif",full.names = TRUE)

# makes a virtual raster layer from the files
vrt(file_list, "altivrt.vrt",overwrite = TRUE)

# import the data from the virtual raster layer
sdf <- rast("altivrt.vrt")

# plotting is very fast (for 190 tiles with 15MB data)
plot(sdf)
```

```{r}
bodenbedeckung <- read_sf("data/MOPUBE_ausschnitt/data/MOPUBE_BBF_ausschnitt.shp")
head(bodenbedeckung)

# remove all columns that are not required
bodenbedeckung <- bodenbedeckung[,-4:-13]

ggplot(bodenbedeckung)+
  geom_sf(aes(fill=BBARTT_BEZ))+
  theme(legend.position = "none")+
  coord_sf(datum = 2056) 

#Clip the Posmo Data to the extent of the groundtype layer. 
boden_subset <- bodenbedeckung[posmo,] # posmo daten 
posmo_subset <- posmo[bodenbedeckung, ] # nur noch polygone angezeigt, die auf der trajectory liegen

# st_is_valid(posmo_cropped, reason=TRUE)
# st_make_valid(posmo_cropped)

# alle Werte im Layer Bodenbedeckung anzeigen, damit dann die passenden für das Mountainbiken definiert werden können.
unique(boden_subset$BBARTT_BEZ)

# check the data
plot(boden_subset[,'BBARTT_BEZ'])
plot(posmo_subset[,'datetime'])

```

We could now see, that there are 23 different groundtypes to describe
what the surface consists of. In order to be able to work faster and to
save memory space, we are now going to spatially join the two layers
bodenbedeckung and posmo, deleting all polygons that do not intersect
with the nodes of the posmo data.

```{r}
# tmap_mode("view")
# tm_shape(boden_subset)+
#   tm_polygons("BBARTT_BEZ")+
#   tm_shape(posmo_subset)+
#   tm_dots()+
#   tm_shape(sdf)+
#   tm_raster()+
#   tmap_options(check.and.fix = TRUE)


plot(sdf)
plot(posmo_subset, add=TRUE)
plot(boden_subset, add=TRUE)
```

This worked well. Looking at the map, we can now see, that the posmo
data is pretty exact and that mostly, the streets are also declared as
streets. The mountainbiking part, however, is simply on "Acker, Wiese,
Weide" terrain. This is not an identifyer for mountainbiking, as it
could also be hiking or other outdoor activities. We will now annotate
the groundtype data to the posmo data and then do the same with the topo
to also add height information to the data in order to achieve filtering
of mountainbiking.

```{r}
posmo_annotated <- st_intersection(posmo_subset, boden_subset)


# suitable for Mountainbiking are the following groundtypes: 
# Acker, Wiese, Weide
# Fels
# (Gartenanlage)
# Geschlossener Wald
# übrige bestockte
# übrige humusierte

mountainbike <- c("Acker, Wiese, Weide", "geschlossener Wald", "Fels", "übrige bestockte", "übrige humusierte")

posmo_annotated2 <- posmo_annotated |>
  filter(BBARTT_BEZ %in% mountainbike) |>  
  mutate(mountainbike_gc = "yes") # mountainbike ground cover

# posmo_annotated2 <- posmo_annotated |>
#   select(BBARTT_BEZ, c("Acker, Wiese, Weide", "Fels", "geschlossener Wald",  "übrige bestockte", "übrige humusierte")) |>
#   mutate(mountainbike = "yes")

unique(posmo_annotated2$BBARTT_BEZ)
```

```{r}
tm_shape(posmo_annotated2)+
  tm_dots("BBARTT_BEZ")
```


### Criteria III: Elevation model

now i will do the same with the terrainmodel

```{r}
posmo_annotated <- posmo_annotated |> 
  mutate(hoehe = extract(sdf, posmo_annotated))

posmo_annotated <- na.omit(posmo_annotated)

posmo_annotated <- posmo_annotated[order(posmo_annotated$datetime, decreasing = FALSE),]

ggplot()+
  geom_point(data= posmo_annotated,
             aes(datetime, hoehe$altivrt, color = BBARTT_BEZ))+
  geom_path(data= posmo_annotated,
            aes(datetime, hoehe$altivrt))
```

in this last visualization we can see very clearly, that the combination
of ground type and elevation will probably make for great criteria to
filter the pattern. In order to filter it from hiking, we will further
add the speed. But first, it needs to be defined that we only want
downhill movements on not paved roads for the further evaluation.

That the height difference is clearly visible, we will compute the
height difference every half minute, which equals 4 points.

```{r}
posmo_annotated <- posmo_annotated |> 
  mutate(
    h_minus1 = (lag(hoehe$altivrt)-hoehe$altivrt),
    h_minus2 = (lag(hoehe$altivrt, 2)-hoehe$altivrt),
    h_plus1 = (lead(hoehe$altivrt) -hoehe$altivrt),
    h_plus2 = (lead(hoehe$altivrt, 2)- hoehe$altivrt)
  )

posmo_annotated <- na.omit(posmo_annotated)

posmo_annotated <- posmo_annotated |>
  rowwise() |>
  mutate(elevmean = mean(c(h_minus1, h_minus2,
                      h_plus1, h_plus2), na.rm=TRUE), 
         elevsum = sum(c(h_minus1, h_minus2,
                      h_plus1, h_plus2), na.rm=TRUE)) |>
  ungroup()

## relevant is summe -> wenn insgesamt bergauf oder bergab?
```

## Combination of criteria

If the sum of the 30 seconds movement block is negative and the
groundtype is either Acker, Wiese, Weide or else geschlossener Wald oder
übrige humusierte, it must be checked with the speed to determine if
these criteria match mountainbiking.

```{r}

```
