---
title: "Rproject_semesterproj"
author: "Lisa Wahlen"
date: "`r Sys.Date()`"
output: html_document
editor_options: 
  markdown: 
    wrap: 72
---

### Libraries

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)


library(readr) # to import tabular data (e.g. csv)
library(dplyr) # to manipulate (tabular) data
library(ggplot2) # to visualize data
library(sf) # to handle spatial vector data
library(terra) # To handle raster data
library(lubridate) # To handle dates and times
library(SimilarityMeasures)
library(RColorBrewer)
library(tmap)

mycolors4 <- brewer.pal(4, "BrBG")
```

### Data

```{r}
posmo <- read_csv("data/posmo_2023-05-18T00_00_00+02_00-2023-05-18T23_59_59+02_00.csv")
```

Evtl. könnte man die Daten auf die AV-Daten legen. So kann man dann alle
Trajectories löschen, welche sich auf einer Strasse befinden.

```{r}
posmo <- st_as_sf(posmo, coords = c("lon_x","lat_y"), crs = 4326) |>
  st_transform(2056)

head(posmo)

posmo_coordinates <- st_coordinates(posmo)

posmo <- cbind(posmo, posmo_coordinates)

posmo_transport <- select(posmo, -c("weekday", "place_name", "user_id"))
posmo <- select(posmo, -c("weekday", "place_name", "transport_mode", "user_id"))
```

### Visualizations

```{r}
# overview datapoints
ggplot(posmo, aes(x=X, y=Y))+
  geom_point()+
  coord_fixed()

# overview transport mode
ggplot(posmo_transport, aes(x=X, y=Y, color=transport_mode))+
  geom_point()+
  coord_fixed()

# overview with map background
tmap_mode("view")
tm_shape(posmo_transport)+
  tm_dots("transport_mode")

```

most of the trajectory is labelled with the transport mode "car". The
loop located in the south of the trajectory shows the part where Lisa
went downhill-biking in Diemtigtal, Allmiried. Most of that section is
labelled as "car", wherease some points are labelled as "walk". We now
want to try to find out some criteria in order to be able to isolate
that downhill biking part and label it as "downhill biking"

## Isolation of mountainbiking pattern

### Criteria I: speed

how to isolate downhill biking pattern?

things to try - setting speed threshold - compare to street map -
compare to elevation model (topography) - segmentation into segments
expressing different sinuosity (Laube, computational movement analysis,
2014)

speed threshold compute movement parameter speed for every fix and a
profile of the movement parameter. (Dodge et al., 2009)

```{r}
head(posmo_transport) # there is an outlier. one point was recorded the night before the tour started.

# remove outlier

posmo_transport <- posmo_transport |> 
  slice(-1)

# calculate distance between points d_plusx for distance
# calculate time between points
# calculate speed
# calculate average speed between four fixes

posmo_transport <- posmo_transport |> 
  mutate(
    d_plus1 = sqrt((lead(X)-X)^2 + (lead(Y)-Y)^2),
    d_plus2 = sqrt((lead(X,2)-X)^2 + (lead(Y,2)-Y)^2), # new offset of 2 in function lead()
    d_minus1 = sqrt((lag(X)-X)^2 + (lag(Y)-Y)^2),
    d_minus2 = sqrt((lag(X,2)-X)^2 + (lag(Y,2)-Y)^2),
    t_plus1 = difftime(lead(datetime), datetime),
    t_plus2 = difftime(lead(datetime, 2), datetime),
    t_minus1 = difftime(lag(datetime), datetime),
    t_minus2 = difftime(lag(datetime, 2), datetime),
    s_plus1 = d_plus1/as.numeric(t_plus1), # s for speed [m/s]
    s_plus2 = d_plus2/as.numeric(t_plus2),
    s_minus1 = d_minus1/-as.numeric(t_minus1),
    s_minus2 = d_minus2/-as.numeric(t_minus2)
  )

# calculate mean speed within window of two neighboring points. not exactly a temporal window because the sampling time varies slightly

posmo_transport <- posmo_transport |>
  rowwise() |>  # so the mean per row is computed
  mutate(
    speedmean = mean(c(s_plus1, s_plus2, s_minus1, s_minus2), na.rm=TRUE)
  ) |> 
  ungroup() # otherwise each row would represent a group, we don't want that

# somehow returns incorrect mean values


# explore what we just did
ggplot(posmo_transport, aes(speedmean))+
  geom_histogram(binwidth = 1)+
  geom_vline(xintercept = mean(posmo_transport$speedmean, na.rm=TRUE)) #no vline because of nan values

range(posmo_transport$speedmean, na.rm=TRUE)

posmo_transport$speedmean[1:50] # how did i get inf values? something divided by 0 returns inf. 0/0 returns NaN
pt <- is.finite(posmo_transport$speedmean) # returns TRUE if it is not one of the values NA, NaN, Inf or -Inf
length(pt[pt == FALSE]) # counts NA, NaN, Inf and -Inf. we have 80 of those

```

Aurelia 19.06.2023

```{r}
# plot points with inf and non numeric value. let's see which points are affected.

inf_filter <- posmo_transport |> 
  filter(!is.finite(speedmean))

tmap_mode("view")
tm_shape(posmo_transport)+
  tm_dots()+
  tm_shape(inf_filter)+
  tm_dots("transport_mode")

# of 2,237 rows only 80 rows are affected (3.58 %). since the +-Inf, NaN and NA values emerge where either the distance or the time between two points is 0, the points in question seem to be static (either temporally or locally). where no speed can be computed, no evaluation regarding a threshold can be taken. 

# new dataset to work with segments
posmo_segments <- posmo_transport |> 
    mutate(static = !is.finite(posmo_transport$speedmean))

# "clean" dataset for visualisation of moving point parameter profile
posmo_transport <- posmo_transport |>
  filter(is.finite(speedmean))

ggplot(posmo_transport, aes(speedmean))+
  geom_histogram(binwidth = 1)+
  geom_vline(xintercept = mean(posmo_transport$speedmean))

tm_shape(posmo_transport)+
  tm_dots("transport_mode")

```

Aurelia 20.06.2023 moving parameter (MP) profile for speed (Dodge et
al., 2009)

```{r}
head(posmo_transport)

ggplot(posmo_transport, aes(x = datetime, y = speedmean))+
  geom_line()+
  geom_hline(yintercept = mean(posmo_transport$speedmean))

```

Segments visualisation

```{r}

ggplot(posmo_segments, aes(X, Y))+
  geom_path()+
  geom_point(aes(color = static))+
  coord_fixed()


```

assign unique IDs to the segments

```{r}
# function for assigning unique ID to the subtrajectories

rle_id <- function(vec) {
    x <- rle(vec)$lengths
    as.factor(rep(seq_along(x), times = x))
}

posmo_segments <- posmo_segments |> 
  mutate(ID=rle_id(static))


```

Visualize the moving segments by colourizing them by segment_ID.

```{r}

posmo_segments <- posmo_segments |> 
  filter(!static) # select rows where the static value is not TRUE to select the moving segments

ggplot(posmo_segments, aes(X, Y))+
  geom_point()+
  geom_path(aes(color=ID))+
  coord_fixed()

tm_shape(posmo_segments)+
  tm_dots("ID")+
  tmap_options(max.categories = 37)

# MP profile
ggplot(posmo_segments, aes(x = datetime, y = speedmean))+
  geom_line(aes(color = ID))+
  geom_hline(yintercept = mean(posmo_transport$speedmean))
```

from the speed profile plot i suppose that the biking pattern starts
around 9 o'clock and ends around half past one. to set a reasonable
speed range or threshold, i want to figure out the average speed of a
bike descent. i try to isolate the bike segments.

```{r}

```

### Criteria II: Ground cover

in a second step, we will try to do the according filtering but by using
a classification according to the DEM and to ground cover.

```{r}
#csv has all download paths from alti3d tiles:  https://www.swisstopo.admin.ch/en/geodata/height/alti3d.html 
# all_tif <- read.csv("data/alti3D_all.csv", header = FALSE)

# terra help https://rspatial.org/spatial-terra/8-rastermanip.html 

#download all files to an empty folder
# for (fi in all_tif$V1){
  # outfile <- basename(fi)
  # print(outfile)
  # download.file(fi, paste0("data/alti3d_05/", outfile),mode = "wb") # mode (binary) very important
# }

# list takes all files with .tif from a folder
file_list <- list.files("data/alti3d_05/",".tif",full.names = TRUE)

# makes a virtual raster layer from the files
vrt(file_list, "altivrt.vrt",overwrite = TRUE)

# import the data from the virtual raster layer
sdf <- rast("altivrt.vrt")

# plotting is very fast (for 190 tiles with 15MB data)
plot(sdf)
```

```{r}
bodenbedeckung <- read_sf("data/MOPUBE/data/MOPUBE_BBF_ausschnitt.shp")

# remove all columns that are not required
bodenbedeckung <- bodenbedeckung[,-4:-13]

#Clip the Posmo Data to the extent of the groundtype layer
boden_subset <- bodenbedeckung[posmo,]
posmo_subset <- posmo[bodenbedeckung, ]

# st_is_valid(posmo_cropped, reason=TRUE)
# st_make_valid(posmo_cropped)

# alle Werte im Layer Bodenbedekcung anzeigen, damit dann die passenden für das Mountainbiken definiert werden können.
unique(boden_subset$BBARTT_BEZ)

# check the data
plot(boden_subset[,'BBARTT_BEZ'])
plot(posmo_subset[,'datetime'])
```

We could now see, that there are 23 different groundtypes to describe
what the surface consists of. In order to be able to work faster and to
save memory space, we are now going to spatially join the two layers
bodenbedeckung and posmo, deleting all polygons that do not intersect
with the nodes of the posmo data.

```{r}
tmap_mode("view")
tm_shape(boden_subset)+
  tm_polygons("BBARTT_BEZ")+
  tm_shape(posmo_subset)+
  tm_dots()+
  tm_shape(sdf)+
  tm_raster()+
  tmap_options(check.and.fix = TRUE)

plot(sdf)
plot(posmo_subset, add=TRUE)
plot(boden_subset, add=TRUE)
```

This worked well. Looking at the map, we can now see, that the posmo
data is pretty exact and that mostly, the streets are also declared as
streets. The mountainbiking part, however, is simply on "Acker, Wiese,
Weide" terrain. This is not an identifyer for mountainbiking, as it
could also be hiking or other outdoor activities. We will now annotate
the groundtype data to the posmo data and then do the same with the topo
to also add height information to the data in order to achieve filtering
of mountainbiking.

```{r}
posmo_annotated <- st_intersection(posmo_subset, boden_subset)
```

### Criteria III: Elevation model

now i will do the same with the terrainmodel

```{r}
posmo_annotated <- posmo_annotated |> 
  mutate(hoehe = extract(sdf, posmo_annotated))

posmo_annotated <- na.omit(posmo_annotated)

posmo_annotated <- posmo_annotated[order(posmo_annotated$datetime, decreasing = FALSE),]

ggplot()+
  geom_point(data= posmo_annotated,
             aes(datetime, hoehe$altivrt, color = BBARTT_BEZ))+
  geom_path(data= posmo_annotated,
            aes(datetime, hoehe$altivrt))
```

in this last visualization we can see very clearly, that the combination
of ground type and elevation will probably make for great criteria to
filter the pattern. In order to filter it from hiking, we will further
add the speed. But first, it needs to be defined that we only want
downhill movements on not paved roads for the further evaluation.

That the height difference is clearly visible, we will compute the
height difference every half minute, which equals 6 points.

```{r}
posmo_annotated <- posmo_annotated |> 
  mutate(
    h_minus1 = (lag(hoehe$altivrt)-hoehe$altivrt),
    h_minus2 = (lag(hoehe$altivrt, 2)-hoehe$altivrt),
    h_minus3 = (lag(hoehe$altivrt, 3)-hoehe$altivrt),
    h_plus1 = (lead(hoehe$altivrt) -hoehe$altivrt),
    h_plus2 = (lead(hoehe$altivrt, 2)- hoehe$altivrt),
    h_plus3 = (lag(hoehe$altivrt, 3)- hoehe$altivrt),
  )

posmo_annotated <- na.omit(posmo_annotated)

posmo_annotated <- format(posmo_annotated, scientific=FALSE)

posmo_annotated <- posmo_annotated |>
  rowwise() |> 
  mutate(elevmean = sum(c(h_minus1, h_minus2, h_minus3,
                      h_plus1, h_plus2, h_plus3), na.rm=TRUE)) |> 
  ungroup()
```

## Combination of criteria
