---
title: "Semesterproject paper"
author: "Aurelia Möri, Lisa Wahlen"
date: "`r Sys.Date()`"
output: html_document
---

Length of report (approx. 15000 char (incl. spaces, incl. References list, excl. Code listing), 20000 char max)

```{r include = FALSE}
# knitr::opts_chunk$set(include = FALSE)
library(wordcountaddin)
library(koRpus)
wordcountaddin:::text_stats()
```

```{r include=FALSE}
# knitr::opts_chunk$set(echo = TRUE)


library(readr) # to import tabular data (e.g. csv)
library(dplyr) # to manipulate (tabular) data
library(ggplot2) # to visualize data
library(sf) # to handle spatial vector data
library(terra) # To handle raster data
library(lubridate) # To handle dates and times
library(SimilarityMeasures)
library(RColorBrewer)
library(tmap)


mycolors4 <- brewer.pal(4, "BrBG")
# BrBG palette: #a6611a, #dfc27d, #80cdc1, #018571
theme_set(theme_light())
```

```{r include=FALSE}
# read Data of the example tour Wiriehorn 05-18

posmo <- read_csv("data/posmo_2023-05-18T00_00_00+02_00-2023-05-18T23_59_59+02_00.csv")

posmo <- st_as_sf(posmo, coords = c("lon_x","lat_y"), crs = 4326) |>
  st_transform(2056)

head(posmo)

posmo_coordinates <- st_coordinates(posmo)
posmo <- cbind(posmo, posmo_coordinates)

head(posmo) # there is an outlier. one point was recorded the night before the tour started.
# remove outlier
posmo <- posmo |> 
  slice(-1)

posmo_transport <- select(posmo, -c("weekday", "place_name", "user_id"))
posmo <- select(posmo, -c("weekday", "place_name", "transport_mode", "user_id"))

tmap_mode("view")
```

```{r include=FALSE} 
# read data for Wiriehorn 05-07 for verification

wiriehorn <- read_csv("data/posmo/mtb_wiriehorn.csv")
head(wiriehorn)
wiriehorn <- st_as_sf(wiriehorn, coords = c("lon_x","lat_y"), crs = 4326) |>
  st_transform(2056)

wiriehorn_coordinates <- st_coordinates(wiriehorn)
wiriehorn <- cbind(wiriehorn, wiriehorn_coordinates)
head(wiriehorn)

# again there is an outlier
wiriehorn <- wiriehorn |> 
  slice(-1)

wiriehorn_transport <- select(wiriehorn, -c("weekday", "place_name", "user_id"))
```

```{r include=FALSE}
# read data of the valbirse tour on 05-06

valbirse <- read_csv("data/posmo/mtb_valbirse.csv")
head(valbirse)
valbirse <- st_as_sf(valbirse, coords = c("lon_x","lat_y"), crs = 4326) |>
  st_transform(2056)

valbirse_coordinates <- st_coordinates(valbirse)
valbirse <- cbind(valbirse, valbirse_coordinates)
head(valbirse)

# again there is an outlier
valbirse <- valbirse |> 
  slice(-1)

valbirse_transport <- select(valbirse, -c("weekday", "place_name", "user_id"))

tmap_mode("view")
```


## Abstract

## Introduction
As mountainbiking is a popular sport in Switzerland (Fischer et al., 2021), both environmental and policy issues arise (Pröbstl-Haider et al., 2018). In order to better understand mountainbiking, this project takes a closer look at the movment of downhill mountainbiking. and be able to talk about these issues on an objective level, it could be of interest to find out more about mountainbiking from a spatial use and movement based perspective. 

movement data of interest. describing movement in GIScience
In this project, we aim at filtering mountainbiking movement patterns in crowdsourced movement data according to specific movement criteria which should only apply to mountianbikers.

In this project, we aim at defining criteria that indicate mountainbiking movement patterns, and try to provide information about the characteristics of mountainbiking. 

what are characteristics of mountainbiking? -> downhill (not uphill), in vegetated areas / trails (not on streets), within a certain speed range. 

In order to filter the mountainbiking pattern, we defined three criteria with which we should be able to have a unique identifier. Those three criteria consist of the speed with which the person is moving, the groundtype on which the trajectory is laying and furthermore the steepness of the trajectory, which needs to be directed downhill. 
Using just one or two of the criteria will always still allow for other activities. Such that a trajectory placed on grass and in forests could still also be a hiker, which also applies to the downhill movement. The Speed of a mountainbiker is hard to define, but it is also not a unique identifier for this movement pattern. 

## Research Question
In this paper we will work on the following two research questions: 
- With which criteria can we filter a downhill oriented mountainbiking movement pattern from Posmo data? 
- Is this combination of criteria applicable to all our data?


## Data and Methods

### Datasets

For the project, I used the movement trajectories that were created by Lisa Wahlen when going mountainbiking. The trajectories were recorded by using the GPS-tracking App posmo (Genossenschaft Posmo Schweiz, 2022) with a sampling rate of 10 seconds. We include 3 mountainbiking tours Lisa completed in the period from the 6th of May to the 18th of May 2023. The tours were recorded in Switzerland in the areas of Wiriehorn (2 tours) and Valbirse (1 tour). 

For the assessment of the criteria that describe the movement pattern of mountain biking, we used the MOPUBE vector dataset providing the land cover types of the canton of Bern (Amt für Geoinformation des Kantons Bern, 2023). In addidtion, we worked with parts of the swiss digital terrain model swissALTI3D (Bundesamt für Landestopografie swisstopo, 2022). This elevation data is provided as raster in a resolution of 0.5 m in grids of 1 km2. We worked with extracts of the datasets, covering the spatial extent of the trajectories. 

We chose the tour Lisa made on the 18th of May in the Wiriehorn area (Wiriehorn 05-18) to define the characteristics of the biking movement pattern and tried to apply the criteria to the tour at Wiriehorn on the 7th of May (Wiriehorn 05-07) and the tour in Valbirse on the 6th of May for verification. The trajectories also included the car ride from Solothurn to the biking location and back. A first visualisation of transport modes of the data revealed that the mountainbiking part of the trajectories were labelled by the posmo app as "Car", "Bus", "Other" - and sometimes almost correctly with "Bike". 

**Wiriehorn trajectory of the 18th of May**  
*Navigating to the south of the trajectory you'll find the mountainbiking part.*
```{r echo=FALSE}
# overview transport mode

tm_shape(posmo_transport)+
  tm_dots("transport_mode")
```

**Wiriehorn trajectory of the 7th of May**  
*Navigating to the south of the trajectory you'll find the mountainbiking part.*
```{r echo=FALSE}

tm_shape(wiriehorn_transport)+
  tm_dots("transport_mode")
```


**Valbirse trajectory of the 6th of May**  
*Navigating to the northeast of the trajectory you'll find the mountainbiking part.*
```{r, echo=FALSE}
tm_shape(valbirse_transport)+
  tm_dots("transport_mode")
```


### Preprocessing

All datasets of the trajectories were first converted to a georeferenced feature. Temporal outliers recorded the night before the trip started were removed, as well as unnecessary or redundant information about Lisa's user ID, the weekday, or the place name.

Preprocessing Bodenbedeckung? in ArcGIS 50m buffer on both sides of the trajectory line

Preprocessing swissALTI3D? polygon extract during acquisition.

### Methods

Three relevant criteria for mountainbiking are speed, downhill movement and ground cover. For all three criteria I assessed the values and ranges fitting the mountainbiking pattern of the Wiriehorn 05-18 tour. Then I applied the criteria to the Wiriehorn 05-06 and the Valbirse tour. 

#### Speed

```{r include=FALSE}
# calculate distance between points d_plusx for distance
# calculate time between points
# calculate speed
# calculate average speed between four fixes

posmo_transport <- posmo_transport |> 
  mutate(
    d_plus1 = sqrt((lead(X)-X)^2 + (lead(Y)-Y)^2),
    d_plus2 = sqrt((lead(X,2)-X)^2 + (lead(Y,2)-Y)^2), # new offset of 2 in function lead()
    d_minus1 = sqrt((lag(X)-X)^2 + (lag(Y)-Y)^2),
    d_minus2 = sqrt((lag(X,2)-X)^2 + (lag(Y,2)-Y)^2),
    t_plus1 = difftime(lead(datetime), datetime),
    t_plus2 = difftime(lead(datetime, 2), datetime),
    t_minus1 = difftime(lag(datetime), datetime),
    t_minus2 = difftime(lag(datetime, 2), datetime),
    s_plus1 = d_plus1/as.numeric(t_plus1), # s for speed [m/s]
    s_plus2 = d_plus2/as.numeric(t_plus2),
    s_minus1 = d_minus1/-as.numeric(t_minus1),
    s_minus2 = d_minus2/-as.numeric(t_minus2)
  )

# calculate mean speed within window of two neighboring points. not exactly a temporal window because the sampling time varies slightly. mean in m/s and km/h.

posmo_transport <- posmo_transport |>
  rowwise() |>  # so the mean per row is computed
  mutate(
    speedmean_ms = mean(c(s_plus1, s_plus2, s_minus1, s_minus2), na.rm=TRUE)
  ) |> 
  ungroup() |> # otherwise each row would represent a group, we don't want that.
  # now  we calculated the mean speed in m/s.
  # calculate mean speed in km per hour. 1 m/s = 3.6 km/h
  mutate(speedmean_kmh = speedmean_ms * 3.6)

```

For every fix on the trajectory I calculated the average speed within the window of four neighboring points, similar to the moving window method of Laube & Purves (2011). The euclidean distance to the two points before and the two points after the fix was calculated and divided by the time difference between the fix and the two points before and after, respectively. Fixes where the speed calculation could not be completed because either the distance or the time between two points was 0, were regarded as static. On the Wiriehorn 05-18 tour, 80 points were labelled as static, and used for segmentation of the trajectory. They were located in places on the trajectory where little movement or a change in transport mode is plausible (e.g. at the bottom and the top of the cable car allmiried, somewhere in the middle of the downhill trail, at train stations in Solothurn).

The extraction of the mountainbiking parts of the trajectory was based on visual assessment of the movement parameter profile (Dodge et al., 2009) for the average speed, and selection of the relevant segments. The selected moving segments included the cable car Lisa took to get uphill to the trail start.
To isolate the cable car and determine the relevant speed range for mounainbiking, the average speed distribution of the selected segments was calculated. The speed values that best characterized the movement of mountain biking were identified through a process of testing.


#### Ground cover type

Intersection of ground cover dataset and the movement points for annotation. 
track annotation with environmental data (Dodge et al., 2013) GlobCover (http://due.esrin.esa.int/page_globcover.php) is from 2009, MOPUBE is updated weekly.
The map 22 ground cover types. Since we assume that mountainbiking takes place on vegetated and/or natural groundcovers, every groundcover type seems suitable except for "Abbau, Deponie", Bahn", "Gebäude", "Strasse, Weg", "Trottoir", "übrige befestigte" and "Verkehrsinsel". we could argue that water bodies are not suitable as well, but depending on their size it is not impossible to encounter such features when biking.

#### Downhill movement

The elevation data was added to the dataset of the Wiriehorn 05-18 trajectory and the difference in elevation to the next two points in the direction of movement was calculated. If within this 20 second window the difference in altidude was negative, it is assumed to be downhill movement.

#### Combination of all criteria

For every point on the trajectory I counted how many of the three criteria applied. 

- try new segmentation by criteria 1) within a window of 60 secs (3 plus,  3 minus), there must be a fix that complies to all three criteria. 2) within a window of 40 secs, the neighbouring 4 fixes must show a sum of matched criteria of at least 6. 

### Limitations
- we should try our code on another mountainbike pattern from Lisa to find whether its applicable to all patterns. 
- visually choosing relevant segments
- quantyfying mountainbiking points, evaluation
- cable car within segments. no segments only including biking
- speed range chosen to exclude cable car.


## Results

### Speed
The average Speed on the Allmiried tour was 37.33548 km/h. The biking segments (including the cable car) showed an average speed of 7.80428 km/h. Since Lisa took the cable car several times on the Allmiried tour, the second and the third quartile of the speed distribution represent the speed of the cable car. In consequence, the first and the fourth quartile of the distribution were chosen to represent mountainbiking, resulting in a range of speed values from 0 to 2.855 and from 10.308 to 26.95490. These criteria also preserve the transport mode "walking".



map of all trajectories 

## Discussion

track annotation with environmental data (Dodge et al., 2013) GlobCover (http://due.esrin.esa.int/page_globcover.php) is from 2009, MOPUBE is updated weekly.



## References

APA 7th edition instructions: <https://libguides.jcu.edu.au/apa/books#s-lg-box-21191622>

Amt für Geoinformation des Kantons Bern. (2023). *Amtliche Vermessung vereinfacht* [Map].  2023https://www.agi.dij.be.ch/de/start/geoportal/geodaten/detail.html?type=geoproduct&code=MOPUBE

Bundesamt für Landestopografie swisstopo. (2022). *swissALTI3D: das hochaufgelöste Terrainmodell der Schweiz* [Map]. https://www.swisstopo.admin.ch/de/geodata/height/alti3d.html

Dodge, S., Weibel, R., & Forootan, E. (2009). Revealing the physics of movement: Com-
paring the similarity of movement characteristics of different types of moving objects.
*Computers, Environment and Urban Systems, 33*(6), 419–434. https://doi.org/10.1016/j.compenvurbsys.2009.07.008

Dodge, S., Bohrer, G., Weinzierl, R., Davidson, S. C., Kays, R., Douglas, D., Cruz, S., Han, J., Brandes, D. & Wikelski, M. (2013). The environmental-data automated track annotation (Env-DATA) system: linking animal tracks with environmental data.
*Movement Ecology, 1*(1), 3. https://doi.org/10.1186/2051-3933-1-3

Fischer, A., Lamprecht, M., & Bürgi, R. (2021). *Mountainbiken in der
Schweiz 2020: Auswertung Mountainbikeland-Befragung 2019
und Sekundäranalyse von «Sport Schweiz 2020»*. Bundesamt für Strassen ASTRA und Stiftung
SchweizMobil. https://www.astra.admin.ch/astra/de/home/themen/langsamverkehr/materialien.html

Genossenschaft Posmo Schweiz. (2022). *Posmo Project App - Tracking für Gruppen*. https://posmo.coop/produkte/posmo-project-tracking-fuer-gruppen

Laube, P. (2014). *Computational Movement Analysis*. Springer Cham. <https://doi.org/10.1007/978-3-319-10268-9>

Laube, P. & Purves, R.(2011) How fast is a cow? Cross-scale Analysis of Movement Data, *Transactions in GIS, 15*(3), 401–418, John Wiley & Sons Ltd, DOI:
10.1111/j.1467-9671.2011.01256.x

***Dodge, S., Laube, P., & Weibel, R. (2012). Movement similarity assessment using symbolic representation of trajectories. *International Journal of Geographical Information Science, 26*(9), 1563-1588. https://doi.org/10.1080/13658816.2011.630003***

Pröbstl-Haider, U., Lund-Durlacher, D., Antonschmidt, H. & Hödl, C. (2018). Mountain bike tourism in Austria and the Alpine region – towards a sustainable model for multi-stakeholder product development. *Journal of Sustainable Tourism, 26*(4), 567-582. https://doi.org/10.1080/09669582.2017.1361428

